Modelling multi-scale state-switching functional data with hidden Markov models

### 2021-05-22 Report/Sections/intro.tex
update reviewer 1, comments 1-4a

Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}. For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time. Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
%, although they ignore sequential dependence between dives. 
%Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
%

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
%, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. 
For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

%In addition, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks. The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change. %To model between-curve dependence in dive profiles, previous studies have categorized killer whale (\textit{Orcinus orca}) dives into different types associated with specified behaviours such as resting, travelling, and foraging \citep{Tennessen:2019b}.
%Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. 
%Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks. The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.

Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. %However, these multilevel models are not appropriate in many biologging applications since they rely on  do not account for a discrete number curve types. 
However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time, which is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.

Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} 
%use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
or with random regression \citep{Rice:2001}. 
%For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves. However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classic FDA techniques.

We aim to identify and describe distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves. To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of
%methodology for the analysis of animal movement 
%statistical ecology
animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, 
%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
but this approach does not account for certain types of temporal correlation.

While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
%
Firstly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. 
Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure. 
%
Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves). 
To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
%
Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales. For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM. \citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing 
daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.

%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence. The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
%and thus extend the methods available for the analysis of functional data. 
We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

>>>>

Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}. For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time. Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. %However, these multilevel models are not appropriate in many biologging applications since they do not account for a discrete number curve types. 
However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time. This is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.

Traditional FDA techniques similarly fall short when modelling complicated within-curve data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} or with random regression \citep{Rice:2001}. However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classic FDA techniques.

The goal of this paper is to identify and describe a finite number of distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves. To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, but this approach does not account for certain types of temporal correlation.

While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
Firstly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves). The fine-scale, within-curve process is of particular interest in this work. To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
Secondly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales. For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM. \citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.

In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence. The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.
